---
title: "Workbook"
author: "Patrick Ballantyne"
date: '2022-09-23'
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(include = FALSE) 
```

## Libraries

Libraries are an essential part of the R workflow, giving you all the necessary tools to complete the analysis. Different people use different libraries for certain aspects of analysis, but these are my favourites:

```{r}
library(tidyverse)
library(sf)
library(tidycensus)
library(tmap)
tmap_mode("view")
```

## Part 1) Explore the data!

Today we will be working with three main types of data:

- Store data, supplied by SafeGraph
- Geodemographic classification, as from paper
- Population data, from the US Census. 

### Store data - SafeGraph places

```{r }
places <- st_read("Data/GA_MSA_pts.gpkg", quiet = TRUE)
```

Look at the attributes:

```{r}
glimpse(places)
```

Let's take a look at the distribution of grocery stores:

```{r}
grocery <- places %>%
  filter(top_category == "Grocery Stores") %>%
  mutate(brands = case_when(brands == "" ~ "Unclassified",
                              TRUE ~ brands))
```

```{r}
tm_shape(grocery) +
  tm_dots(col = "brands", size = 0.05)
```

Lots of 'independent' grocery stores, but also large numbers of big brands such as LIDL, Circle K and Whole Foods. 

### US GEODEMOGRAPHIC CLASSIFICATION 

```{r}
geodemo <- st_read("Data/GA_GD.gpkg", quiet = TRUE)
```

```{r}
glimpse(geodemo)
```

The TractGroup column is the one we are interested in; every census tract in the US is assigned to one of 10 groups, based on the geodemographic characteristics of the area in question. 

```{r}
tm_shape(geodemo) +
  tm_fill(col = "TractGroup", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.5)
```

### POPULATION DATA

We are going to be using a fantastic package called 'Tidycensus' which is used to directly query data from the US Census Bureau API.

```{r}
census_api_key("e0f3cdfa0e1e13ab3a4d9dc32c67984d0be5523b")
```

```{r message= FALSE}
pop <- get_decennial(geography = "Tract", variables = "P001001",
                     state = "GA", year = 2010)
```

```{r}
glimpse(pop)
```

Here, the value column is the one we are interested in; total population in each census tract. 

```{r}
gd_pop <- merge(geodemo, pop, by = "GEOID")
```

```{r}
glimpse(gd_pop)
```

Now that we have the population data converted to a spatial form, let's take a look at the population distribution of Atlanta MSA:

```{r}
tm_shape(gd_pop) +
  tm_fill(col = "value", alpha = 0.5, style = "jenks") +
  tm_borders(col = "black", lwd = 0.5)
```

## PART 2 - CATCHMENT ANALYSIS



## PART 3 - CATCHMENT PROFILING


